Inequality in socioemotional skills: a cross-cohort comparison

### 2018-09-22 codes/R/paper/paper_jpube.Rnw
revise EFA, add loadings table

\usepackage{authblk}
\title{Inequality in non-cognitive skills: a cross-cohort comparison}
\author[1,2]{Orazio Attanasio}
\author[1,2]{Richard Blundell}
\author[1,2]{Gabriella Conti}
\author[1,2]{Giacomo Mason}
\affil[1]{\emph{Institute for Fiscal Studies}}
\affil[1]{\emph{UCL Department of Economics}}
%\institute{UCL and IFS}
\date{\today}

@@

@@

<<AUXFUN>>=
# load function that gets parameters
@

@@

<<MEANTABLE_PRINT>>=
# function to print mean table
printmeantab <- function(tab) {
  print(xtable(tab,
               align="L{0cm}C{.04\\textwidth}C{.05\\textwidth}C{.05\\textwidth}L{.35\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}L{.35\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
               digits = 1
  ), 
  size="\\small",
  include.rownames=F,
  include.colnames=F,
  booktabs = T,
  add.to.row = list(pos = c(list(0)),
                    command = c(
                      # header
                      paste(" & & &
                            \\multicolumn{4}{c}{\\textbf{BCS}} &
                            \\multicolumn{4}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){4-7} \\cmidrule(lr){8-11} 
                            Item & Factor & Num. Categ. &
                            Wording & Cert. Appl. (\\%) & Smtm. Appl. (\\%) & Appl. (\\%) &
                            Wording & Cert. True (\\%) & Smwt. True (\\%) & True (\\%) \\\\ \\midrule
                            "
                            , sep="")
                    )),
  sanitize.text.function = function(x){x},
  sanitize.rownames.function = function(x){""}, # to not print row numbers
  tabular.environment = "tabular", width = ".8\\textwidth",
  floating = FALSE
  )
}

@@

\begin{table}[]
\centering\setstretch{1.2} \footnotesize
\caption{Parameterisations for measurement invariance}\label{tab:invparam}
\begin{tabular}{L{.18\textwidth}L{.45\textwidth}L{.3\textwidth}}
\toprule
\textbf{Invariance level}     & \textbf{Description} & \textbf{Restrictions} \\ \midrule \\
Configural (WE{\textTheta})   
    & \bi[label=$\cdot$] \item Minimally restrictive model for identification \ei
    & For all groups: $\begin{array}{|l} \text{diag}(\bm{\Phi}) = \bm{I} \\
                      \bm{\kappa}=\bm{0} \\
                      \bm{\nu}=\bm{0} \\
                      \text{diag}(\bm{\Psi})=\bm{I} \end{array}$
                      \\ \\
Threshold invariance
    & \bi[label=$\cdot$] \item Restricts thresholds $\tau$ to be equal across groups
                  \item Statistically equivalent to configural (when measures have 3 categories or less)\ei
    & $\tau_{1,ci} = \tau_{1,c^\prime i}$ for all items, $\forall c, c^\prime$
                      \newline $\tau_{2,ci} = \tau_{2,c^\prime i}$ for non-binary items, $\forall c, c^\prime$
                      \newline For all groups: $\begin{array}{|l} \text{diag}(\bm{\Phi}) = \bm{I} \\ 
                                                \bm{\kappa}=\bm{0} \end{array}$
                      \newline For ref. group $A$: $\;\begin{array}{|l} \bm{\nu}_A=\bm{0} \\
                                                  \text{diag}(\bm{\Sigma}_A)=\bm{I} \end{array}$
                   \\ \\
Threshold and Loading invariance
    & \bi[label=$\cdot$] \item Restricts thresholds $\tau$ and loadings $\lambda$ to be equal across groups
                  \item Allows comparison of latent factor variances \ei
    & $\tau_{1,ci} = \tau_{1,c^\prime i}$ for all items, $\forall c, c^\prime$
                      \newline $\tau_{2,ci} = \tau_{2,c^\prime i}$ for non-binary items, $\forall c, c^\prime$
                      \newline $\lambda_{ci} = \lambda_{c^\prime i}$ for all items, $\forall c, c^\prime$
                      \newline For all groups: $\bm{\kappa}=\bm{0}$
                      \newline For ref. group $A$: $\;\begin{array}{|l} \bm{\nu}_A=\bm{0} \\
                                                  \text{diag}(\bm{\Sigma}_A)=\bm{I} \\
                                                  \text{diag}(\bm{\Phi}_A) = \bm{I} \end{array}$ 
                   \\ \\
Threshold, Loading, and Intercept invariance
    & \bi[label=$\cdot$] \item Restricts thresholds $\tau$ and loadings $\lambda$ to be equal across groups
                  \item Restricts intercepts $\nu$ to zero in both groups
                  \item Allows comparison of latent factor variances \emph{and} means \ei
    & $\tau_{1,ci} = \tau_{1,c^\prime i}$ for all items, $\forall c, c^\prime$
                      \newline $\tau_{2,ci} = \tau_{2,c^\prime i}$ for non-binary items, $\forall c, c^\prime$
                      \newline $\lambda_{ci} = \lambda_{c^\prime i}$ for all items, $\forall c, c^\prime$
                      \newline For all groups: $\bm{\nu}=\bm{0}$
                      \newline For ref. group $A$: $\;\begin{array}{|l} \bm{\kappa}_A=\bm{0} \\
                                                  \text{diag}(\bm{\Sigma}_A)=\bm{I} \\
                                                  \text{diag}(\bm{\Phi}_A) = \bm{I} \end{array}$ 
\\ \\ \bottomrule
\end{tabular}
\end{table}

% ADD PARTIAL INVARIANCE

@@

@@

\begin{landscape}
\begin{table}[]
\caption{Item wording and prevalence, by cohort -- females}\label{tab:meantab_f}
\centering\setstretch{1.2} \footnotesize
<<MEANTAB_F, dependson="MEANTABLE", results='asis'>>=
require(xtable)
table <- meantab[[2]] # females
table <- table[,!names(table) %in% c("bcs.num", "mcs.num")] # drop if unneeded
printmeantab(table)
@
\end{table}
\end{landscape}

\begin{landscape}
\begin{table}[]
\caption{Item wording and prevalence, by cohort -- males}\label{tab:meantab_m}
\centering\setstretch{1.2} \footnotesize
<<MEANTAB_M, dependson="MEANTABLE", results='asis'>>=
require(xtable)
table <- meantab[[1]] # males
table <- table[,!names(table) %in% c("bcs.num", "mcs.num")] # drop if unneeded
printmeantab(table)
@
\end{table}
\end{landscape}

>>>>

\usepackage{authblk}
\title{Inequality in socioemotional skills: a cross-cohort comparison}
\author[1,2]{Orazio Attanasio}
\author[1,2]{Richard Blundell}
\author[1,2]{Gabriella Conti}
\author[1,2]{Giacomo Mason}
\affil[1]{\emph{Institute for Fiscal Studies}}
\affil[1]{\emph{UCL Department of Economics}}
%\institute{UCL and IFS}
\date{\today}

@@

\begin{abstract}
abstract here
\end{abstract}

% We are grateful to The Centre for Longitudinal Studies, UCL Institute of Education for the use of these data and to the UK Data Archive and UK Data Service for making them available. However, they bear no responsibility for the analysis or interpretation of these data.

@@

<<AUXFUN>>=
# load auxiliary functions
@

@@

<<BCSEFA, dependson='LOAD_DATA'>>=
@

<<BCSEFARES, dependson='BCSEFA'>>=
@

<<MCSEFA, dependson='LOAD_DATA'>>=
@

<<MCSEFARES, dependson='MCSEFA'>>=
@

@@

@@

\clearpage
\begin{table}[]
\centering\setstretch{1.2} \footnotesize
\caption{Parameterisations for measurement invariance}\label{tab:invparam}
\begin{tabular}{L{.18\textwidth}L{.45\textwidth}L{.3\textwidth}}
\toprule
\textbf{Invariance level}     & \textbf{Description} & \textbf{Restrictions} \\ \midrule \\
Configural (WE{\textTheta})   
    & \bi[label=$\cdot$] \item Minimally restrictive model for identification \ei
    & For all groups: $\begin{array}{|l} \text{diag}(\bm{\Phi}) = \bm{I} \\
                      \bm{\kappa}=\bm{0} \\
                      \bm{\nu}=\bm{0} \\
                      \text{diag}(\bm{\Psi})=\bm{I} \end{array}$
                      \\ \\
Threshold invariance
    & \bi[label=$\cdot$] \item Restricts thresholds $\tau$ to be equal across groups
                  \item Statistically equivalent to configural (when measures have 3 categories or less)\ei
    & $\tau_{1,ci} = \tau_{1,c^\prime i}$ for all items, $\forall c, c^\prime$
                      \newline $\tau_{2,ci} = \tau_{2,c^\prime i}$ for non-binary items, $\forall c, c^\prime$
                      \newline For all groups: $\begin{array}{|l} \text{diag}(\bm{\Phi}) = \bm{I} \\ 
                                                \bm{\kappa}=\bm{0} \end{array}$
                      \newline For ref. group $A$: $\;\begin{array}{|l} \bm{\nu}_A=\bm{0} \\
                                                  \text{diag}(\bm{\Sigma}_A)=\bm{I} \end{array}$
                   \\ \\
Threshold and Loading invariance
    & \bi[label=$\cdot$] \item Restricts thresholds $\tau$ and loadings $\lambda$ to be equal across groups
                  \item Allows comparison of latent factor variances \ei
    & $\tau_{1,ci} = \tau_{1,c^\prime i}$ for all items, $\forall c, c^\prime$
                      \newline $\tau_{2,ci} = \tau_{2,c^\prime i}$ for non-binary items, $\forall c, c^\prime$
                      \newline $\lambda_{ci} = \lambda_{c^\prime i}$ for all items, $\forall c, c^\prime$
                      \newline For all groups: $\bm{\kappa}=\bm{0}$
                      \newline For ref. group $A$: $\;\begin{array}{|l} \bm{\nu}_A=\bm{0} \\
                                                  \text{diag}(\bm{\Sigma}_A)=\bm{I} \\
                                                  \text{diag}(\bm{\Phi}_A) = \bm{I} \end{array}$ 
                   \\ \\
Threshold, Loading, and Intercept invariance
    & \bi[label=$\cdot$] \item Restricts thresholds $\tau$ and loadings $\lambda$ to be equal across groups
                  \item Restricts intercepts $\nu$ to zero in both groups
                  \item Allows comparison of latent factor variances \emph{and} means \ei
    & $\tau_{1,ci} = \tau_{1,c^\prime i}$ for all items, $\forall c, c^\prime$
                      \newline $\tau_{2,ci} = \tau_{2,c^\prime i}$ for non-binary items, $\forall c, c^\prime$
                      \newline $\lambda_{ci} = \lambda_{c^\prime i}$ for all items, $\forall c, c^\prime$
                      \newline For all groups: $\bm{\nu}=\bm{0}$
                      \newline For ref. group $A$: $\;\begin{array}{|l} \bm{\kappa}_A=\bm{0} \\
                                                  \text{diag}(\bm{\Sigma}_A)=\bm{I} \\
                                                  \text{diag}(\bm{\Phi}_A) = \bm{I} \end{array}$ 
\\ \\ \bottomrule
\end{tabular}
\end{table}

@@

\begin{landscape}
\begin{table}[]
\caption{Item prevalence, by cohort and gender}\label{tab:meantab}
\centering\setstretch{1.2} \footnotesize
<<MEANTAB, dependson="MEANTABLE", results='asis'>>=
require(xtable)
table <- meantab # females
table <- table[,!names(table) %in% c("bcs.text", "bcs.num", "mcs.text", "mcs.num")] # drop if unneeded

  print(xtable(table,
               align="L{0cm}C{.04\\textwidth}C{.05\\textwidth}C{.05\\textwidth}L{.12\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
               digits = 1
  ), 
  include.rownames=F,
  include.colnames=F,
  booktabs = T,
  add.to.row = list(pos = c(list(0)),
                    command = c(
                      # header
                      paste(" & & & &
                            \\multicolumn{6}{c}{\\textbf{Males}} & \\multicolumn{6}{c}{\\textbf{Females}}
                            \\\\ \\cmidrule(lr){5-10} \\cmidrule(lr){11-16} 
                            & & & & 
                            \\multicolumn{3}{c}{\\textbf{BCS}} & \\multicolumn{3}{c}{\\textbf{MCS}}
                            & \\multicolumn{3}{c}{\\textbf{BCS}} & \\multicolumn{3}{c}{\\textbf{MCS}}
                            \\\\ \\cmidrule(lr){5-7} \\cmidrule(lr){8-10} \\cmidrule(lr){11-13} \\cmidrule(lr){14-16} 
                            & & & &
                            Cert. Appl. & Smtm. Appl. & Appl. &  Cert. True & Smwt. True & True &
                            Cert. Appl. & Smtm. Appl. & Appl. &  Cert. True & Smwt. True & True 
                            \\\\
                            Itm. & Factor & Cat. & Title & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) \\\\
                            "
                            , sep="")
                    )),
  sanitize.text.function = function(x){x},
  sanitize.rownames.function = function(x){""}, # to not print row numbers
  tabular.environment = "tabular",
  floating = FALSE
  )

@
\justify {\scriptsize \emph{Notes}: The table shows the prevalence by gender and cohort for each item of our novel subscale. \emph{Itm.} is item number. \emph{Factor} is the latent construct to which the item loads -- EXT is Externalising skills, INT is Internalising skills. \emph{Cat.} is the number of categories in which the item is coded -- 2 denotes a binary item (applies/does not apply) and 3 denotes a 3-category item. \emph{Title} is a short label for the item. \emph{Cert. / Smtm. Appl.} =  Certainly / sometimes applies. \emph{Cert. / Smwt. True} = Certainly / somewhat true.
\par}
\end{table}
\end{landscape}

\begin{table}
\caption{Suggested number of factors to retain}\label{tab:numfac}
\centering\setstretch{1.2}\footnotesize
<<NUMFACTAB, dependson='MCSEFARES', results='asis'>>=
# combine results
efatab <- data.frame(cbind(efatab.bcs,efatab.mcs))

efatab %>%
  tibble::rownames_to_column() %>%
  kable("latex", escape = F, booktabs = T, linesep = "", align=c('l','c','c','c','c','c','c'),
        col.names = c("Approach", "All", "Males", "Females", "All", "Males", "Females")) %>%
  add_header_above(c(" " = 1,  "BCS (1970)" = 3, "MCS (2000/1)" = 3)) %>%
  column_spec(2:7, width = "3em")

@
\justify {\scriptsize \emph{Notes}: The table compares the optimal number of factors suggested by different aprroaches for our novel scale; scree test based approaches (optimal coordinates, acceleration factor -- \citealp{Raiche2013}), parallel analysis \citep{Horn1965}, Kaiser's eigenvalue rule \citep{Kaiser1960}, Very Simple Structure (VSS, \citealp{Revelle1979}), Velicer Minimum Average Partial test (MAP, \citealp{Velicer1976}).
\par}
\end{table}

\begin{table}
\caption{Loadings from exploratory factor analysis}\label{tab:load}
\centering\setstretch{1.2}\footnotesize
<<EFALOADTAB, dependson='MCSEFARES', results='asis'>>=
# combine results
library(kableExtra)
loadtab <- cbind(meantab$num, meantab$title, data.frame(cbind(bcsfa$loadings, mcsfa$loadings)))
colnames(loadtab) <- c("item", "title", "bcs1", "bcs2", "mcs1", "mcs2")

loadtab %>%
  mutate_at(vars(matches("cs")), round, 3) %>%
  mutate(
    bcs1 = cell_spec(bcs1, "latex", bold = ifelse(abs(bcs1) > .25, TRUE, FALSE)),
    bcs2 = cell_spec(bcs2, "latex", bold = ifelse(abs(bcs2) > .25, TRUE, FALSE)),
    mcs1 = cell_spec(mcs1, "latex", bold = ifelse(abs(mcs1) > .25, TRUE, FALSE)),
    mcs2 = cell_spec(mcs2, "latex", bold = ifelse(abs(mcs2) > .25, TRUE, FALSE))
  ) %>%
  kable("latex", escape = F, booktabs = T, linesep = "", align=c('c','l','c','c','c'),
        col.names = c("Item", "Title", "Factor 1 (EXT)", "Factor 2 (INT)", "Factor 1 (EXT)", "Factor 2 (INT)")) %>%
  add_header_above(c(" " = 2,  "BCS (1970)" = 2, "MCS (2000/1)" = 2))

@
\justify {\scriptsize \emph{Notes}: The table displays the factor loadings obtained from exploratory factor analysis (EFA) on our novel scale, separately by cohort. The EFA is based on the decomposition of the polychoric correlation matrix, and uses weighted least squares and oblimin rotation.
\par}
\end{table}
